---
title: "A Colorful First Project"
subtitle: about Stats and Floyd-Steinberg Dithering
author: "Justin Kraaijenbrink"
date: "`r format(Sys.time(), '%d %B, %Y')`"
graphics: yes
output: 
  html_document:
      fig_caption: yes
      keep_tex: yes
header-includes: 
- \usepackage{graphicx}
- \usepackage{float}
- \usepackage{placeins}
- \usepackage{amsmath}
- \newcommand{\argmin}{\operatornamewithlimits{argmin}}
- \newcommand{\argmax}{\operatornamewithlimits{argmax}}
---

```{r label = setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  dev.args = list(bg = 'transparent'),
  fig.align = "center", 
  fig.height = 5, 
  cache = TRUE,
  echo = TRUE,
  eval = FALSE
)
library("tidyverse")
library("png")
```

Read in image:
```{r, eval = TRUE}
quokka <- png::readPNG(source = "0_img/quokka.png", native = FALSE) 
quokka <- quokka[, , -4] # first three matrices only
```

# 1 About Colors: A Picture = A Data set.

## 1.1
```{r, eval = TRUE}
pixelcols <- quokka[600, 10, ]

pixel <- tibble(Color = c("Red", "Green", "Blue"), Pixel_Intensity = pixelcols)

pixel
```

This pixel will be black! 


## 1.2 
```{r, eval = TRUE}
quokka_df <- tibble(row = rep(1:nrow(quokka), ncol(quokka)),
            col = rep(1:ncol(quokka), each = nrow(quokka)),
            red = as.vector(quokka[, , 1]),
            green = as.vector(quokka[, , 2]),
            blue = as.vector(quokka[, , 3]),
            rgb_color = rgb(quokka[, , 1], quokka[, , 2], quokka[, , 3]))
```


## 1.3 
Check if the number of values in character variable `rgb_color` is equal to the
number of observed combinations of `red`, `green` and `blue`. 
```{r, eval = TRUE}
unique_combs <- quokka_df %>%
  select("red", "green", "blue") %>% 
  unite(., "all_combs", sep = "-")%>% 
  unique() %>% 
  nrow()

unique_cols <- quokka_df %>%
  select("rgb_color") %>% 
  unique() %>% 
  nrow()

unique_combs == unique_cols
```

The `TRUE` in the output hopefully speaks for itself!

## 1.4
Create a raster to plot picture in R:
```{r, eval = TRUE}
quokka_img <- matrix(quokka_df$rgb_color, 
                     nrow = nrow(quokka), 
                     ncol = ncol(quokka)) %>% as.raster()
```

```{r, eval = TRUE}
png("0_img/quokka_img.png", width = ncol(quokka), height = nrow(quokka), units = "px")
par(mar = c(0, 0, 0, 0))
plot(quokka_img)
dev.off()
```

\begin{center}
\includegraphics[height = 5cm]{0_img/quokka_img.png}
\end{center}

Such a cute animal!

\clearpage

# 2 A Further Understanding of the RGB Space

## 2.1
Program own RGB-function:
```{r, eval = TRUE}
CreateRGBColors <- function(red, green, blue, maxValue = 1){
  
  colors <- c(red, green, blue)
  
  if(!all(round(colors) %in% 0:1) && maxValue == 1){
    stop("Color intensities must be in [0, 1]")
  } else if(!all(round(colors) %in% 0:255) && maxValue == 255){
    stop("Color intensities must be in [0, 255]")
  } 
  
  if(maxValue == 1) {
    colors <- 255 * colors
  }
  
  RoundToUneven <- function(x) {
    for(i in 1:length(x)){
      if(trunc(x[i]) %% 2 == 0 && x[i] - trunc(x[i]) == 0.5){
        x[i] <- ceiling(x[i])
      } else {
        x[i] <- round(x[i])
      }
    }
    return(x)
  }
  
  colors <- RoundToUneven(colors)

  hexadecimal <- c(0:9, LETTERS[1:6])
  hdm2columns <- expand.grid(hexadecimal, hexadecimal)
  channel <- paste0(hdm2columns[, 2], hdm2columns[, 1], sep = "")
  
  color.ind <- colors + 1
  
  rgbcolor <- paste0(c("#", channel[color.ind]), collapse = "")
  
  return(rgbcolor)
}
```

I decided to design the function such that the output resembles the output from the `rgb()` function (that displays an error when the color intensities are not in [0, 1], but `maxColorValue` is). 

Show that both functions give the same output:
```{r}
identical(CreateRGBColors(0.5, 0.3, 0.7, maxValue = 1), 
          rgb(0.5, 0.3, 0.7, maxColorValue = 1))
```

## 2.2
```{r, eval = TRUE}
CreatePalette <- function(K){
  
  if(K %in% c(8, 64, 512, 4096, 32768) == FALSE & K %in% c(3, 6, 9, 12, 15) == FALSE){
    stop("Input K must be a one of these numbers: 
         (8, 64, 512, 4096, 32768) or (3, 6, 9, 12, 15)")
  }
  
  if(K %in% c(3, 6, 9, 12, 15) == FALSE & log2(K) %% 3 != 0){
    stop("Your input must satisfy the following condition: log2(K) %% 3 == 0")
  } 

  if(K %in% c(8, 64, 512, 4096, 32768)){
    K = K^(1/3)
  }
  
  if(K %in% c(3, 6, 9, 12, 15)){
    K = 2^(K/3)
  }
  
  color_combs <- expand.grid(seq(0, 255, 255/(K - 1)), 
                             seq(0, 255, 255/(K - 1)), 
                             seq(0, 255, 255/(K - 1)))
  
  colnames(color_combs) <- c("Red", "Green", "Blue")
  
  all_colors <- apply(color_combs, 1, function(x){
    red = x[1]
    green = x[2]
    blue = x[3]
    
    col <- CreateRGBColors(red, green, blue, maxValue = 255)
    })

  out <- list(cols = all_colors,
              dat = color_combs)
  
  return(out)
}
```

## 2.3

### 2.3a
I decided to put the code into a function, so that we can easily adjust things such as the number of bits.
Input for the function is:

1. an image (array)
2. number of bits you want the picture to be compressed to
3. maxValue (indicates whether the intensities are in [0, 1] or in [0, 255])

The function output is a new array (`dim = c(382, 513, 3)`) with the compressed intensities. 

```{r, eval = TRUE}
FindClosestPaletteColor <- function(picture, n_bit, maxValue = 1){
  if(maxValue != 1 & maxValue != 255){
    stop("maxValue should be either 1 or 255")
  }
  
  if(maxValue == 1){
    picture <- picture * 255
  }
  
  col_palette <- CreatePalette(n_bit)$dat
  
  closestpixel <- array(dim = dim(picture))
  
  for(i in 1:nrow(picture)){
    for(j in 1:ncol(picture)){
      pixel <- picture[i, j, ]
      ind <- which.min(rowSums(t(pixel - t(col_palette))^2))
      
      closestpixel[i, j, ] <- as.numeric(col_palette[ind, ])
    }
  }
  
  return(closestpixel)
}
```

### 2.3b 

```{r, eval = TRUE}
quokka_03bit <- FindClosestPaletteColor(n_bit = 3, 
                                        picture = quokka, 
                                        maxValue = 1) / 255
```

```{r, eval = TRUE}
writePNG(quokka_03bit, "0_img/quokka_03bit.png")
```


# 3 Floyd-Steinberg dithering algorithm

## 3.1 

First, I create a Loss-function which takes as input a pixel (vector with length 3) and the number of bits you want to compress your picture to.
This function can be used in the DitherImage-function.

```{r}
LossFunction <- function(pixel, col_palette = col_palette){
  ind <- which.min(rowSums(t(pixel - t(col_palette))^2))
  
  newpixel <- as.numeric(col_palette[ind, ])
  
  return(newpixel)
}
```

Creating the `DitherImage`-function, with the image-array and number of bits as inputs:

```{r}
DitherImage <- function(picture, n_bit){
  
  pixels <- picture
  
  quant_error <- array(dim = dim(picture))
  
  col_palette <- CreatePalette(n_bit)$dat/255
  
  for(i in 1:(nrow(pixels) - 1)){
    for(j in 1:(ncol(pixels) - 1)){
        oldpixel <- pixels[i, j, ] 
        newpixel <- LossFunction(oldpixel, col_palette)
        pixels[i, j, ] <- newpixel
        quant_error[i, j, ] <- oldpixel - newpixel
        
        pixels[i    , j + 1, ] <-  pixels[i    , j + 1, ] + quant_error[i, j, ] * 7 / 16
        pixels[i + 1, j + 1, ] <-  pixels[i + 1, j + 1, ] + quant_error[i, j, ] * 1 / 16
        pixels[i + 1, j    , ] <-  pixels[i + 1, j    , ] + quant_error[i, j, ] * 5 / 16
        pixels[i + 1, j - 1, ] <-  pixels[i + 1, j - 1, ] + quant_error[i, j, ] * 3 / 16
      }
  }

  col_oldpixels <- pixels[, ncol(pixels), ]
  col_newpixels <- matrix(nrow = nrow(col_oldpixels), ncol = ncol(col_oldpixels))
  for(i in 1:nrow(col_oldpixels)){
    col_newpixels[i, ] <- LossFunction(col_oldpixels[i, ], col_palette)
  }
  pixels[, ncol(pixels), ] <- col_newpixels
  quant_error[, ncol(pixels), ] <- col_oldpixels - col_newpixels

  row_oldpixels <- pixels[nrow(pixels), , ]
  row_newpixels <- matrix(nrow = nrow(row_oldpixels), ncol = ncol(row_oldpixels))
  for(i in 1:nrow(row_oldpixels)){
    row_newpixels[i, ] <- LossFunction(row_oldpixels[i, ], col_palette)
  }
  pixels[nrow(pixels), , ] <- row_newpixels
  quant_error[nrow(pixels), , ] <- row_oldpixels - row_newpixels
  
  out <- list(New_Colors = pixels,
              Errors = 255 * quant_error,
              Loss = sum((255 * quant_error)^2))
  
  return(out)
}
```

To see whether my results are (approximately) the same as the modelanswers for the 3-bit image:

```{r}
dithered_03 <- DitherImage(quokka, n_bit = 3)
writePNG(dithered_03$New_Colors, "0_img/quokka_dithered_03.png")
```

